
---
title: "Examining Trends in Pediatric Healthcare Utilization"
author: "Laura Robles-Torres"
date: "February 18, 2025"
output:
  html_document:
    code_folding: hide
subtitle: "Data Analyst Candidate Assignment: CCHSR"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Part 1: Trends in utilization of various healthcare services among children between 2015 and 2023

*The code below includes initial library loading, data importing, and reshaping for analysis.*

```{r import sheet 1, results='hide', message=FALSE, warning=FALSE, code_folding="hide"}
#Load necessary libraries
library(tidyverse)
library(tidyr)
library(readxl)
library(dbplyr)
library(stringr)
library(ggplot2)

#Import Sheet 1 data
utilization_data =
    readxl::read_excel("data.xlsx", sheet = "Sheet1")

head(utilization_data)
```

#### Healthcare utilization by insurance payer

```{r payor groups, code_folding="hide",  warning=FALSE}
#Clean data for payor groups 
payor_data = utilization_data |>
  mutate(
    payor = ifelse(category %in% c("Medicaid", "Private Insurance"), category, NA) 
  ) |> 
  fill(payor, .direction = "down") |> 
  filter(!category %in% c("Medicaid", "Private Insurance")) |> # Remove payer rows that are empty 
  slice(1:8) #Use only data reflecting utilization by payer

# Reshape data into long format
payor_data |>
  pivot_longer(cols = `2015`:`2023`, names_to = "year", values_to = "utilization") |>
  mutate(year = as.numeric(year)) |>
  select(year, payor, category, utilization) -> payor_data
```


```{r summary tables for payor,  message=FALSE, warning=FALSE,}
# Table 1: Mean utilization by insurance type
payor_table <- payor_data |> 
  group_by(payor, category) |>  
  rename(Payor=payor) |>
  summarise(mean_utilization = mean(utilization, na.rm = TRUE), .groups = "drop") |>
  mutate(across(where(is.numeric), ~ format(ceiling(.), big.mark = ","))) |>
  pivot_wider(names_from = category, values_from = mean_utilization)

# Display tables
knitr::kable(payor_table, caption = "Mean Utilization by Insurance Type")
```

Plotting trends by service over time 
```{r plotting payor trends,  warning=FALSE}
#Plot utilization of each service by payor
ggplot(payor_data, aes(x = year, y = utilization, color = payor)) +
  geom_line() +
  facet_wrap(~ category, scales = "fixed") +
  theme_minimal() +
  labs(title = "Healthcare Utilization Trends by Payor (2015-2023)", x = "Year", y = "Utilization", color="Insurance Payor") +
  theme(
    plot.title = element_text(hjust = 0.5))  # Center the title
```

#### Healthcare utilization by race/ethnicity
```{r race/ethnicity groups, code_folding="hide", warning=FALSE}
#Clean data for racial/ethnic group data 
race_data = 
  utilization_data |>
  mutate(
    race = ifelse(category %in% c("Hispanic", "Non-Hispanic Black", "Non-Hispanic White"), category, NA)
  ) |> 
  fill(race, .direction = "down") |>  # Fill race downwards
  filter(!category %in% c("Hispanic", "Non-Hispanic Black", "Non-Hispanic White")) # Remove race rows that are empty

# Reshape data into long format
race_data |>
  pivot_longer(cols = `2015`:`2023`, names_to = "year", values_to = "utilization")|>
  mutate(year = as.numeric(year)) |> 
  filter(!is.na(utilization)) |> 
  filter(!is.na(race)) |> #Filter out missing values if any 
  slice_tail(n=126) -> race_data #Use only data reflecting utilization by race
```


```{r  race/ethnicity groups table, message=FALSE, warning=FALSE,}
# Table 2: Mean utilization by race/ethnicity
race_table <- race_data |> 
  group_by(race, category) |>  
  summarise(mean_utilization = mean(utilization, na.rm = TRUE), .groups = "drop") |>  
  pivot_wider(names_from = category, values_from = mean_utilization) |>
  mutate(across(where(is.numeric), ~ format(ceiling(.), big.mark = ","))) |>
  rename(Race=race) 

knitr::kable(race_table, caption = "Mean Utilization by Race/Ethnicity")

```

```{r plotting race/ethnicity trends, code_folding="hide",  warning=FALSE}
#Plot utilization of each service by race/ethnicity
ggplot(race_data, aes(x = year, y = utilization, color = race)) +
  geom_line() +
  facet_wrap(~ category, scales = "fixed") +
  theme_minimal() +
  labs(title = "Healthcare Utilization Trends by Race/Ethnicity (2015-2023)", x = "Year", y = "Utilization", color = "Race/Ethnicity") +
    theme(
    plot.title = element_text(hjust = 0.5))  # Center the title
```

### Part 2: Demographic trends among children who underwent surgery for congenital heart disease in NY state in 2022

```{r, part 2 import & clean,  warning=FALSE}
options(scipen = 999)  # Turn off scientific notation globally so data correctly imports from Excel sheet 

#Import sheet 2 (patient data)
sheet2 <- readxl::read_excel("data.xlsx", sheet = "Sheet2") 

sheet2$tract <- as.numeric(sheet2$tract)  # Convert 'tract' column to numeric to facilitate join 

#Import sheet 3 (census tract data)
sheet3 <- readxl::read_excel("data.xlsx", sheet = "Sheet3") |>
          janitor::clean_names(case = "snake") |> 
          separate(geo_id, sep="S", into = c("geo_id", "tract")) |> #Extract 'tract' from 'geo_id' variable for join 
          mutate(tract = as.numeric(tract)) #Ensure tract is numeric as well 

 clean_sheet3 <- sheet3 |> #Rename variables of interest for ease
  rename(pop_6_m=b27003_003e, pop_18_m = b27003_006e, pop_6_f = b27003_031e, pop_18_f = b27003_034e) |>
  rename(public_6_m = b27003_004e, public_18_m = b27003_007e, public_6_f = b27003_032e, public_18_f = b27003_035e, pop_density = populationdensitypersquarem) |> 
  select(-geo_id)
 
 clean_sheet3 <-clean_sheet3 |> #Select only variables of interest 
    select(tract, pop_6_m, pop_18_m,  pop_6_f, pop_18_f, public_6_m, 
         public_18_m, public_6_f, public_18_f, pop_density) 
```

```{r link data,  message=FALSE, warning=FALSE,}
#Link two sheets by tract
linked_data = 
  inner_join(sheet2,clean_sheet3, by=c("tract"))
```

```{r,  message=FALSE, warning=FALSE,}
#Descriptive statistics of sample 
#Age
age_stats <- linked_data |> 
  summarise(
    Mean_Age = mean(age, na.rm = TRUE),
    Median_Age = median(age, na.rm = TRUE),
    SD_Age = sd(age, na.rm = TRUE),
    Min_Age = min(age, na.rm = TRUE),
    Max_Age = max(age, na.rm = TRUE),
  )

# Descriptive statistics for sex and race/ethnicity
sex_stats <- linked_data |>
  group_by(sex) |>
  summarise(
    Count = n(),
    Proportion = n() / nrow(linked_data)  # Proportion of each sex
  )

race_stats <- linked_data |>
  group_by(race) |>
  summarise(
    Count = n(),
    Proportion = n() / nrow(linked_data)  # Proportion of each race/ethnicity
  )

```

#### Proportion of people ≤18 years old who are on public insurance for each patient’s census tract:

*The table below is a sample to show final calculations.*

```{r, calculating proportion, warning=FALSE, results='hide'}
linked_data |>
  mutate(total_pop = pop_6_m + pop_18_m + pop_6_f + pop_18_f) |>
  mutate(public_pop = public_6_m + public_18_m + public_6_f + public_18_f) |>
  mutate(proportion_public = ifelse(total_pop == 0, 0, public_pop / total_pop)) -> proportion_public_per_tract  
# Replace NaN with 0 if total_pop is 0 

sample_proportion_public <-
  proportion_public_per_tract |>
  mutate(proportion_public = round(proportion_public, 2)) |>
  select(tract, total_pop, public_pop, proportion_public) |>
  slice_sample(n=10)

knitr::kable(sample_proportion_public) #Sample of 10 random census tracts to show calculations

#List all unique tracts (no duplicates) and their respective calculated proportions
unique_tracts <- proportion_public_per_tract |>
  distinct(tract, .keep_all = TRUE)

#See how many unique tracts are missing population density data 
unique_tracts[complete.cases(unique_tracts), ]
unique_tracts[is.na(unique_tracts$pop_density), ] #1,026 tracts  do not have population density data 

#How many tracts have >90% children on public insurance?
num_tracts_above_90 <- unique_tracts|>  
  filter(proportion_public > 0.90) |>  
  nrow()

#How many tracts have <10% children on public insurance?
num_tracts_below_10 <- unique_tracts|>  
  filter(proportion_public < 0.10) |>  
  nrow()

print(num_tracts_above_90)
print(num_tracts_below_10)
```

```{r density analysis, message=FALSE, warning=FALSE,}
unique_tracts |>
  filter(!is.na(pop_density)) |>
  mutate(proportion_public = round(proportion_public, 2)) -> prop_public_tract_complete

#1,026 cases do not have population density variable data and were excluded from this analysis 

#Stratification
prop_public_tract_stratified <- prop_public_tract_complete |> 
  mutate(density_category = case_when(
    pop_density < quantile(pop_density, 0.33, na.rm = TRUE) ~ "Low",
    pop_density < quantile(pop_density, 0.66, na.rm = TRUE) ~ "Medium",
    TRUE ~ "High"
  ))

#Summary of average proportion by density 
density_summary <- prop_public_tract_stratified |> 
  group_by(density_category) |> 
  summarise(mean_proportion = mean(proportion_public, na.rm = TRUE))

knitr::kable(density_summary)

#Visualizing relationship between density and public insurance proportion, transformed on log 
ggplot(unique_tracts , aes(x = pop_density, y = proportion_public)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    x = "Population Density (per square mile)", 
    y = "Proportion of Children on Public Insurance"
  ) +
  scale_x_log10() +  # Apply log scale to x-axis
  theme_minimal()
```

```{r checking distribution of data, message=FALSE, warning=FALSE,}
#Visualize the distribution of proportions by tracts
ggplot(unique_tracts, aes(x = proportion_public)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Public Insurance Enrollment",
       x = "Proportion of Children on Public Insurance",
       y = "Count of Census Tracts") +
  theme_minimal()

#Median proportion of children on public insurance across all tracts.  
median_prop <- unique_tracts |>
  filter(!is.na(proportion_public)) |>
  summarise(
    median_public_insurance = median(proportion_public, na.rm = TRUE) #Calculated median due to skewness of data 
  )

# Print the result
print(median_prop) 
```

#### Patterns by public insurance and race/ethnicity

The sample is relatively evenly distributed across racial/ethnic groups.

```{r patterns of public insurance coverage by race/ethnicity and public insurance,  warning=FALSE}
#How many children of each racial/ethnic group are in our dataset?
linked_data |>
  group_by(race) |>
  summarize(n_obs=n()) |>
    arrange(desc(n_obs)) -> grouped_race #we have a pretty balanced sample

knitr::kable(grouped_race, caption = "Number of children grouped by race/ethnicity")
```

##### Do certain racial/ethnic groups tend to have a higher proportion of children of public insurance?  

```{r patterns between public insurance and race/ethnicity, warning=FALSE}
#Proportions by race
proportion_public_per_tract |> 
  filter(!is.na(proportion_public)) |>  # Remove rows with NA or zero pop_density
  group_by(race) |>
  summarize(
    median_prop = median(proportion_public)) -> prop_by_race

knitr::kable(prop_by_race,  caption = "Weighted Mean Proportion of Children on Public Health Insurance by Race/Ethnicity")
```

```{r visualize,  warning=FALSE}
ggplot(proportion_public_per_tract, aes(x = race, y = proportion_public, fill = race)) +
  geom_boxplot() +
  stat_summary(fun = median, geom = "text", aes(label = round(..y.., 3)), 
               position = position_dodge(width = 0.75), vjust = -0.5) +
  labs(title = "Median Proportion of Children ion Public Insurance by Race/Ethnicity",
       x = "Race/Ethnicity",
       y = "Proportion on Public Insurance") +
    scale_fill_discrete(name = "Race/Ethnicity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


